Predicting the 2024 U.S. Presidential Election: Trump’s Likely Victory Based on Polling Analysis

### 2024-11-03 paper/paper.qmd
committing changes to git

---
title: "Predicting a Trump Victory: A State-Level Analysis of the 2024 US Presidential Election Using Bayesian Modeling"
subtitle: "State-by-State Insights on Voter Dynamics Favoring Trump Over Harris in Key Battlgrounds"
author: 
  - Jimin Lee
  - Sarah Ding
  - Xiyan Chen
thanks: "Code and data are available at: [https://github.com/jamiejiminlee/2024_US_Elections.git](https://github.com/jamiejiminlee/2024_US_Elections.git)."
date: today
date-format: long
abstract: "This paper presents a predictive model for the 2024 United States Presidential Election using state-level polling data. By aggregating high-quality polls, we predict the likely winner between Donald Trump and Kamala Harris in each state. A logistic regression model estimates the probability of Trump winning, with polling percentages as predictor variables. The model’s results highlight key battleground states and offer insights into voter dynamics, while also reflecting on the limitations of polling data in forecasting political outcomes. Our findings suggest that Trump holds an adantage in critical swing states, positinoning him as the likely winner of the election."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

## Model Summary Results
```{r}
#| echo: false
#| eval: true
#| label: tbl-fixed-effects-summary
#| tbl-cap: "Fixed Effects Summary for Trump and Harris Models."
#| warning: false

# Load necessary libraries
library(dplyr)
library(broom)        # For the tidy function
library(knitr)
library(kableExtra)

#| echo: false
#| eval: true
#| label: tbl-fixed-effects-summary
#| tbl-cap: "Fixed Effects Summary for Trump and Harris Models."
#| warning: false

# Load necessary libraries
library(dplyr)
library(broom)        # For the tidy function
library(knitr)
library(kableExtra)

# Summarize fixed effects for Trump model
trump_summary <- tidy(trump_model, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Estimate = estimate,
    `Std. Error` = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  mutate(term = gsub("[^[:alnum:] ]", "", term))  # Remove non-alphanumeric characters

# Summarize fixed effects for Harris model
harris_summary <- tidy(harris_model, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Estimate = estimate,
    `Std. Error` = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  mutate(term = gsub("[^[:alnum:] ]", "", term))  # Remove non-alphanumeric characters

# Combine summaries into one table
model_summary_table <- full_join(trump_summary, harris_summary, by = "term", suffix = c(" Trump", " Harris"))

# Display the table with kable
kable(model_summary_table, caption = "Fixed Effects Summary for Trump and Harris Models") %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 8)

@@

@@

@@

```
## Election Day Predictions

@@

@@

# Predict probability of Trump winning by comparing posterior predictions for each state
state_predictions <- clean_poll_data %>%
  distinct(state) %>%
  mutate(end_date = election_day, end_date_num = as.numeric(election_day - min(clean_poll_data$end_date)))

# Calculate probability of Trump winning in each state
state_predictions <- state_predictions %>%
  rowwise() %>%
  mutate(
    prob_trump_wins = mean(posterior_predict(trump_model, newdata = cur_data()) >
                           posterior_predict(harris_model, newdata = cur_data()))
  )

# Plot probability of winning by state
ggplot(state_predictions, aes(x = reorder(state, prob_trump_wins), y = prob_trump_wins, fill = prob_trump_wins > 0.5)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue"), labels = c("Harris", "Trump")) +
  labs(title = "Probability of Trump Winning by State on Election Day", x = "State", y = "Probability Trump Wins") +
  theme_minimal() +
  theme(legend.position = "none")

@@

>>>>

---
title: "Predicting the 2024 U.S. Presidential Election: Trump’s Likely Victory Based on Polling Analysis"
subtitle: "A Bayesian Approach Reveals Strong Support for Donald Trump Across Key States, Highlighting Critical Trends in Voter Preferences Leading Up to Election Day"
author: 
  - Jimin Lee
  - Sarah Ding
  - Xiyan Chen
thanks: "Code and data are available at: [https://github.com/jamiejiminlee/2024_US_Elections.git](https://github.com/jamiejiminlee/2024_US_Elections.git)."
date: today
date-format: long
abstract: "This paper presents a predictive model for the 2024 United States Presidential Election using state-level polling data. By aggregating high-quality polls, we predict the likely winner between Donald Trump and Kamala Harris in each state. A logistic regression model estimates the probability of Trump winning, with polling percentages as predictor variables. The model’s results highlight key battleground states and offer insights into voter dynamics, while also reflecting on the limitations of polling data in forecasting political outcomes. Our findings suggest that Trump holds an adantage in critical swing states, positinoning him as the likely winner of the election."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

## Model Summary - Harris Model
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Harris Model."
#| warning: false

# Summarize fixed effects for Harris model
harris_summary <- tidy(harris_model, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Estimate = estimate,
    `Std. Error` = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  mutate(term = gsub("[^[:alnum:] ]", "", term))  # Clean up term names

# Display the Harris model summary table
knitr::kable(harris_summary) %>%
  kable_styling(full_width = FALSE, position = "center")

@@

EDIT - explain model summary table

## Model summary - Trump Model
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false

# Summarize fixed effects for Trump model
trump_summary <- tidy(trump_model, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Estimate = estimate,
    `Std. Error` = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  mutate(term = gsub("[^[:alnum:] ]", "", term))  # Clean up term names

# Display the Trump model summary table
knitr::kable(trump_summary) %>%
  kable_styling(full_width = FALSE, position = "center")
```

EDIT - explain model summary table

@@

```
EDIT - explain results on graph

@@

```
EDIT - explain results on graph

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

# Predict probability of Trump and Harris winning by comparing posterior predictions for each state
state_predictions <- clean_poll_data %>%
  distinct(state) %>%
  mutate(end_date = election_day, end_date_num = as.numeric(election_day - min(clean_poll_data$end_date)))

# Calculate probabilities of Trump and Harris winning in each state
state_predictions <- state_predictions %>%
  rowwise() %>%
  mutate(
    prob_trump_wins = mean(posterior_predict(trump_model, newdata = cur_data()) > 
                           posterior_predict(harris_model, newdata = cur_data())),
    prob_harris_wins = mean(posterior_predict(harris_model, newdata = cur_data()) > 
                            posterior_predict(trump_model, newdata = cur_data()))
  )

# Prepare the state predictions for the table
state_summary_table <- state_predictions %>%
  select(state, prob_trump_wins, prob_harris_wins) %>%
  mutate(
    prob_trump_wins = round(prob_trump_wins * 100, 2),  # Convert to percentage and round
    prob_harris_wins = round(prob_harris_wins * 100, 2)  # Convert to percentage and round
  )

# Display state predictions i n table
knitr::kable(state_summary_table) %>%
  kable_styling(full_width = FALSE, position = "center")

EDIT - explain table results

@@

# Reshape the data for plotting
state_predictions_long <- state_predictions %>%
  select(state, prob_trump_wins, prob_harris_wins) %>%
  pivot_longer(cols = starts_with("prob"), 
               names_to = "Candidate", 
               values_to = "Probability") %>%
  mutate(Candidate = recode(Candidate, 
                            prob_trump_wins = "Trump", 
                            prob_harris_wins = "Harris"))

# Plot double bar graph of predicted winning percentages by state
ggplot(state_predictions_long, aes(x = reorder(state, Probability), y = Probability * 100, fill = Candidate)) +
  geom_col(position = "dodge") +  # Use dodge for side-by-side bars
  coord_flip() +  # Flip coordinates for better readability
  scale_fill_manual(values = c("Trump" = "red", "Harris" = "blue")) +
  labs(title = "Predicted Winning Percentages by State on Election Day", 
       x = "State", 
       y = "Winning Probability (%)") +
  theme_minimal() +
  theme(legend.position = "right")
```
EDIT - explain graph results

@@

EDIT - explain map results
